Quantifying Stability and Change in Personal Culture Using Panel Data

### 2023-09-24 manuscript.Rmd
included Fig 3 again; moved endnotes to bottom (still need to tweak that); tweaks on intro and beginning of discussion.

\begin{center}
<about here>
Figure 1: Proportions of explained variance in items of personal culture; based on predictions at participants' wave mid-point.
\end{center}

@@

Figure 2 plots the results from an ordinary least squares regression of proportion of systematic variance attributable to intrapersonal change as a function of question, panel, and sample features. These models also include fixed effects for panels (not shown), so coefficients reflect the association between a particular sample or question characteristic within a panel.[^paneleffects]

[^paneleffects]: KK NOTE: Can we summarize these panel fixed effects at all?

\begin{center}
<about here>
Figure 2: Proportion of explained variance attributable to intrapersonal change; based on predictions at participants' wave mid-point; supressed intercept model; survey indicators not shown.
\end{center}

@@

[^missing]: We estimate the models presented in [earlier figure] using the full sample, but only use respondents who have information on their highest degree to generate the results presented in figures [these figures]. In total, ### respondents (%%% percent) reported no degree information.

@@

\begin{center}
<about here>
Figure 3: Proportions of intrapersonal change of explained variance by college degree; based on predictions at participants' wave mid-point.
\end{center}

There is a clear pattern in [figure above]: for more than 80 percent of GSS items explored here, intrapersonal change is a larger component of systematic variance for people without a college degree than for people with a college degree (they have negative values in the figure). While most of these differences are small in absolute terms (less than 2 percentage points), several are greater than 5 percentage points. Given the distribution observed in Figure [the earlier one] showing that the systematic variance attributable to intrapersonal change averages [average across all questions], a 5 percentage point difference between groups is quite substantial.

To more clearly illustrate some of these differences, we highlight eight questions designed to tap general political dispositions: partisan identification (Democrat vs. Republican) and ideological identification (liberal vs. conservative) on seven-point scales; four questions about the government's role in improving the condition of the poor, paying people's medical bills, giving special treatment to blacks, and doing things that private businesses could do, measured on five-point scales; a question about whether the government should do more to reduce income differences, measured on a seven-point scale; and one question about whether blacks should be given preferences in hiring, measured on a five-point scale. Estimates for the proportion of systematic variance attributable to intrapersonal change on these eight questions, for both education groups, are presented in Figure XXX.

On all eight questions presented in Figure XXX, intrapersonal change explains less systematic variance for people with a college degree than for people without a college degree. There are also large differences across questions on the amount of systematic variance attributable to intrapersonal change for both groups. For example, on the question of whether the government should try to solve more problems or leave those problems to be solved by private businesses ("government do more or less"), less than one percent of the systematic variance is attributable to intrapersonal change for both groups. In other words, while people might vacillate on this question at random (37 percent of variance is residual for this question), there is functionally no evidence that people make systematic changes of opinion on this issue during the GSS panel.

@@

This paper set out intervene in the debate on whether people change their personal culture -- their attitudes, beliefs, values, practices, and dispositions -- as they move through life. Instead of falling back on a "tournament of models" approach [@lersch2023: p. 228] to call victory for one or the other possible answer to this question, this paper seeks to refocus the debate to the theoretically more productive question of _how_ initial differences between and durable changes within people collectively explain the observed variation in personal culture. To do so, it proposes a new measure that quantifies the amount of systematic variance that is attributable to either interpersonal differences at baseline or intrapersonal change.

Applying this measure to all datasets previously discussed in this debate revealed a consistent pattern. Nearly all questions on personal culture show evidence of intrapersonal change over time, with some items showing noteably high intrapersonal change compared to interpersonal differences. For questions about factors that affect one’s life satisfaction or views on government benefits, it seems plausible that differences in adult experiences predominantly account for the observed differences between people.

@@

More important than adjudicating past debates, however, the approach presented in this paper offers offers the possibility to pose new theoretical questions regarding when change or stability in personal culture predominates and its significance in explaining cultural variances within a population. To examplify this potential, we detailed how the proportion in cultural differences that intrapersonal change can explain not only varies substantially by survey item but also by individual characteristics such as their education. Intrapersonal change plays almost no part in explaining sentiments regarding the government's economic role but is more defining for ideological and partisan identities. Intriguingly, it is more important in explaining cultural variation among people without a college compared to college graduates in the U.S.. While age and other factors related to college completion might confound this pattern, it suggests that college may solidify personal culture to an extent that renders the effect of other adult experiences on political dispositions for graduates less influential relative to individuals without such a college experience.

@@

@@

\newpage
```{r fig3, out.width = "450px", fig.cap="Proportions of intrapersonal change of explained variance by college degree; based on predictions at participants' wave mid-point."}
knitr::include_graphics("fig/FIG_college2.png")
```

>>>>

\begin{center}
<about here>
\end{center}

Figure 1: Proportions of explained variance in items of personal culture; based on predictions at participants' wave mid-point.

@@

Figure 2 plots the results from an ordinary least squares regression of proportion of systematic variance attributable to intrapersonal change as a function of question, panel, and sample features. These models also include fixed effects for panels (not shown), so coefficients reflect the association between a particular sample or question characteristic within a panel.[KK NOTE: CAN WE SUMMARIZE THESE PANEL FIXED EFFECTS AT ALL? AE: WE COULD SHOW THAT THEY DON'T CAPTURE TOPIC EFFECTS BEYOND OUR CLASSIFICATION IF WE INCLUDE THOSE / THAT THEY POSSIBLY CAPTURE UNMEASURED COUNTRY/CONTEXT DIFFERENCES OR DIFFERENCES IN SURVEY METHODOLOGY?]

\begin{center}
<about here>
\end{center}
Figure 2: Proportion of explained variance attributable to intrapersonal change; based on predictions at participants' wave mid-point; supressed intercept model; survey indicators not shown.

@@

@@

\begin{center}
<about here>
\end{center}
Figure 3: Difference between share of intrapersonal change and share of inter-personal differences in expalined variance; based on predictions at participants' wave mid-point

There is a clear pattern in Figure 4: for more than 80 percent of GSS items explored here, intrapersonal change is a larger component of systematic variance for people without a college degree than for people with a college degree (they have negative values in the figure). While most of these differences are small in absolute terms (less than 2 percentage points), several are greater than 5 percentage points. Given the distribution observed in Figure [the earlier one] showing that the systematic variance attributable to intrapersonal change averages [average across all questions], a 5 percentage point difference between groups is quite substantial.

To more clearly illustrate some of these differences, we highlight eight questions designed to tap general political dispositions: partisan identification (Democrat vs. Republican) and ideological identification (liberal vs. conservative) on seven-point scales; four questions about the government's role in improving the condition of the poor, paying people's medical bills, giving special treatment to blacks, and doing things that private businesses could do, measured on five-point scales; a question about whether the government should do more to reduce income differences, measured on a seven-point scale; and one question about whether blacks should be given preferences in hiring, measured on a five-point scale. Estimates for the proportion of systematic variance attributable to intrapersonal change on these eight questions, for both education groups, are presented in Figure 4.

\begin{center}
<about here>
\end{center}
Figure 4: Proportions of intrapersonal change of explained variance by college degree; based on predictions at participants' wave mid-point.

On all eight questions presented in Figure 4, intrapersonal change explains less systematic variance for people with a college degree than for people without a college degree. There are also large differences across questions on the amount of systematic variance attributable to intrapersonal change for both groups. For example, on the question of whether the government should try to solve more problems or leave those problems to be solved by private businesses ("government do more or less"), less than one percent of the systematic variance is attributable to intrapersonal change for both groups. In other words, while people might vacillate on this question at random (37 percent of variance is residual for this question), there is functionally no evidence that people make systematic changes of opinion on this issue during the GSS panel.

@@

This paper set out intervene in the debate on whether people change their personal culture -- their attitudes, beliefs, values, practices, and dispositions -- as they move through life. Instead of falling back on a "tournament of models" approach [@lersch2023: p. 228] to call victory for one or the other possible answer to this question, it seeks to refocus the debate to the theoretically more productive issue of _how_ initial differences between and durable changes within people collectively explain the variation in personal culture that is observed. To do so, it proposes a new measure that quantifies the amount of systematic variance that is attributable to either interpersonal differences at baseline or intrapersonal change.

Applying this measure to all datasets previously discussed in this debate revealed a consistent pattern. Nearly all questions on personal culture show evidence of durable, intrapersonal change over time, with some items showing noteably high intrapersonal change compared to interpersonal differences. For questions about factors that affect one’s life satisfaction or views on government benefits, it seems plausible that differences in adult experiences predominantly account for the observed differences between people.

@@

More important than adjudicating past debates, however, the approach presented in this paper unlocks new opportunities to ask when change or stability in personal culture predominates and about their role in explaining cultural variation within a population. To exemplify this, we detailed how the proportion in cultural differences that intrapersonal change can explain not only varies substantially by survey item but also by individual characteristics such as their education. Intrapersonal change plays almost no part in explaining sentiments regarding the government's economic role but is more defining for ideological and partisan identities. Intriguingly, it is more important in explaining cultural variation among non-college graduates compared to college graduates in the U.S.. While age and other factors related to college completion might confound this pattern, this suggests that experiencing college may solidify personal culture to an extent that renders the effect of other adult experiences on political dispositions for graduates less influential relative to individuals without a college experience.

@@

# Endnotes

[^missing]: We estimate the models presented in Figures 1 and 2 using the full sample, but only use respondents who have information on their highest degree to generate the results presented in Figures 3 and 4. In total, ### respondents (%%% percent) reported no degree information.

@@

\newpage
```{r fig3, out.width = "450px", fig.cap="Difference between share of intrapersonal change and share of inter-personal differences in expalined variance; based on predictions at participants' wave mid-point."}
knitr::include_graphics("fig/FIG_college1.png")
```

\newpage
```{r fig4, out.width = "450px", fig.cap="Proportions of intrapersonal change of explained variance by college degree; based on predictions at participants' wave mid-point."}
knitr::include_graphics("fig/FIG_college2.png")
```
